Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-03-01 article-models-tt.tex
did all the proofs of the extension on terms subsection

\begin{lem}
When we substitute an extended term $\ctxext{f}{g}$ of $\ctxext{P}{Q}$ by a term
$x$ of $A$, the term that we get is $\ctxext{\subst{x}{f}}{\subst{x}{g}}$.
More precisely, the following inference rules are valid:
\begin{align*}
& \inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}\quad\jterm{\ctxext{\Gamma}{A}}{\subst{f}{Q}}{g}}{\jtermeq{\Gamma}{\ctxext{\subst{x}{P}}{\subst{x}{Q}}}{\subst{x}{\ctxext{f}{g}}}{\ctxext{\subst{x}{f}}{\subst{x}{g}}}}\\
& \inference{\jterm{\ctxext{{\Gamma}{A}}{P}}{Q}{g}\quad\jterm{\ctxext{{\Gamma}{A}}{P}}{\subst{g}{R}}{t}}{\jtermeq{\ctxext{\Gamma}{\subst{x}{P}}}{\ctxext{\subst{x}{Q}}{\subst{x}{R}}}{\subst{x}{\ctxext{g}{t}}}{\ctxext{\subst{x}{g}}{\subst{x}{t}}}}
\end{align*}
\end{lem}

@@

>>>>

\begin{proof}
Consider $\jterm{\ctxext{\Gamma}{B}}{Q}{g}$ and $\jterm{\ctxext{\Gamma}{B}}{\subst{g}{R}}{t}$.
Then we have the judgmental equalities
\begin{align*}
\ctxwk{A}{\ctxext{g}{t}}
& \jdeq \ctxwk{A}{\subst{t}{{g}{\idfunc[\ctxext{Q}{R}]}}}\\
& \jdeq \subst{\ctxwk{A}{t}}{\ctxwk{A}{\subst{g}{\idfunc[\ctxext{Q}{R}]}}}\\
& \jdeq \subst{\ctxwk{A}{t}}{{\ctxwk{A}{g}}{\ctxwk{A}{\idfunc[\ctxext{Q}{R}]}}}\\
& \jdeq \subst{\ctxwk{A}{t}}{{\ctxwk{A}{g}}{\idfunc[\ctxwk{A}{\ctxext{Q}{R}}]}}\\
& \jdeq \subst{\ctxwk{A}{t}}{{\ctxwk{A}{g}}{\idfunc[\ctxext{\ctxwk{A}{Q}}{\ctxwk{A}{R}}]}}\\
& \jdeq \ctxext{\ctxwk{A}{g}}{\ctxwk{A}{t}}
\end{align*}
\end{proof}

\begin{lem}
When we substitute an extended term $\ctxext{f}{g}$ of $\ctxext{P}{Q}$ by a term
$x$ of $A$, the term that we get is $\ctxext{\subst{x}{f}}{\subst{x}{g}}$.
More precisely, the following inference rules are valid:
\begin{align*}
& \inference{\jterm{\ctxext{{\Gamma}{A}}{P}}{Q}{g}\quad\jterm{\ctxext{{\Gamma}{A}}{P}}{\subst{g}{R}}{t}}{\jtermeq{\ctxext{\Gamma}{\subst{x}{P}}}{\ctxext{\subst{x}{Q}}{\subst{x}{R}}}{\subst{x}{\ctxext{g}{t}}}{\ctxext{\subst{x}{g}}{\subst{x}{t}}}}
\end{align*}
\end{lem}

\begin{proof}
Consider $\jterm{\ctxext{\Gamma}{B}}{Q}{g}$ and $\jterm{\ctxext{\Gamma}{B}}{\subst{g}{R}}{t}$.
Then we have the judgmental equalities
\begin{align*}
\subst{x}{\ctxext{g}{t}}
& \jdeq \subst{x}{{t}{{g}{\idfunc[\ctxext{Q}{R}]}}}\\
& \jdeq \subst{{x}{t}}{{x}{{g}{\idfunc[\ctxext{Q}{R}]}}}\\
& \jdeq \subst{{x}{t}}{{{x}{g}}{{x}{\idfunc[\ctxext{Q}{R}]}}}\\
& \jdeq \subst{{x}{t}}{{{x}{g}}{\idfunc[\subst{x}{\ctxext{Q}{R}}]}}\\
& \jdeq \subst{{x}{t}}{{{x}{g}}{\idfunc[\ctxext{\subst{x}{Q}}{\subst{x}{R}}]}}\\
& \jdeq \ctxext{\subst{x}{g}}{\subst{x}{t}}.
\end{align*}
\end{proof}

@@

Pretty type theory is pretty as an ugly little duckling. At first, it is actually uglier than
structural type theory because one has to keep track of a variable nobody is
actually interested in. The situation improves when we introduce the operators
of extension and weakening, because from that point onwards one can draw notational
advantages from having the variable around.

Thus we make eight kinds of judgments in our type theory: ``$\Gamma$ is a context'',
``$A(i)$ is a family of contexts over $\Gamma$'', ``$A$ is a type in context $\Gamma$''
and ``$x$ is a term of the family $A$ of contexts over $\Gamma$''. The other four
judgments are for judgmental equality.

\begin{align*}
\jvctx*{\Gamma} & \jvctxeq*{\Gamma}{\Gamma'}\\
\jvfam*{i}{\Gamma}{A} & \jvfameq*{i}{\Gamma}{A}{B}\\
\jvtype*{i}{\Gamma}{A} & \jvtypeeq*{i}{\Gamma}{A}{B}\\
\jvterm*{i}{\Gamma}{A}{x} & \jvtermeq*{i}{\Gamma}{A}{x}{y}.
\end{align*}

\begin{small}
\begin{align*}
& \inference{\jvtype{i}{\Gamma}{A}}{\jvfam{i}{\Gamma}{A}} & & \inference{\jvtypeeq{i}{\Gamma}{A}{B}}{\jvfameq{i}{\Gamma}{A}{B}}\\
& \inference{\jvtype{i}{\Gamma}{A}\quad\jvfameq{i}{\Gamma}{A}{B}}{\jvtype{i}{\Gamma}{B}}
& & \inference{\jvtype{i}{\Gamma}{A}\quad\jvfameq{i}{\Gamma}{A}{B}}{\jvtypeeq{i}{\Gamma}{A}{B}}
\end{align*}
\end{small}

\subsection{The basic rules for judgmental equality}
The rules for judgmental equality establish that it is an equivalence relation
in all three cases (contexts, types and terms).
\bgroup\small
\begin{align*}
& \inference{\jvctx{\Gamma}}{\jvctxeq{\Gamma}{\Gamma}} 
& & \inference{\jvctxeq{\Gamma}{\Delta}}{\jvctxeq{\Delta}{\Gamma}} 
& & \inference{\jvctxeq{\Gamma}{\Delta}\qquad\jvctxeq{\Delta}{\greek{E}}}{\jvctxeq{\Gamma}{\greek{E}}}\\
& \inference{\jvfam{i}{\Gamma}{A}}{\jvfameq{i}{\Gamma}{A}{A}} 
& & \inference{\jvfameq{i}{\Gamma}{A}{B}}{\jvfameq{i}{\Gamma}{B}{A}}
& & \inference{\jvfameq{i}{\Gamma}{A}{B}\qquad\jvfameq{i}{\Gamma}{B}{C}}{\jvfameq{i}{\Gamma}{A}{C}}\\
& \inference{\jvterm{i}{\Gamma}{A}{x}}{\jvtermeq{i}{\Gamma}{A}{x}{x}}
& & \inference{\jvtermeq{i}{\Gamma}{A}{x}{y}}{\jvtermeq{i}{\Gamma}{A}{y}{x}}
& & \inference{\jvtermeq{i}{\Gamma}{A}{x}{y}\qquad\jvtermeq{i}{\Gamma}{A}{y}{z}}{\jvtermeq{i}{\Gamma}{A}{x}{z}}
\end{align*}
\egroup

\begin{align*}
& \inference{\jvctxeq{\Gamma}{\Delta}\qquad\jvfam{i}{\Gamma}{A}}{\jvfam{i}{\Delta}{A}}
& & \inference{\jvctxeq{\Gamma}{\Delta}\qquad\jvfameq{i}{\Gamma}{A}{B}}{\jvfameq{i}{\Delta}{A}{B}}\\
& \inference{\jvctxeq{\Gamma}{\Delta}\qquad\jvterm{i}{\Gamma}{A}{x}}{\jvterm{i}{\Delta}{A}{x}}
& & \inference{\jvctxeq{\Gamma}{\Delta}\qquad\jvtermeq{i}{\Gamma}{A}{x}{y}}{\jvtermeq{i}{\Delta}{A}{x}{y}}\\
& \inference{\jvfameq{i}{\Gamma}{A}{B}\qquad \jvterm{i}{\Gamma}{A}{x}}{\jvterm{i}{\Gamma}{B}{x}}
& & \inference{\jvfameq{i}{\Gamma}{A}{B}\qquad\jvtermeq{i}{\Gamma}{A}{x}{y}}{\jvtermeq{i}{\Gamma}{B}{x}{y}}
\end{align*}

\begin{align}
& \inference{}{\jctx{\emptyc}}\\
& \inference{\jctx{\Gamma}}{\jvfam{i}{\Gamma}{\emptyf[\Gamma]}}\\
& \inference{\jctx{\Gamma}}{\jvterm{i}{\Gamma}{\emptyf[\Gamma]}{\emptytm[\Gamma]}}\\
& \inference{\jvterm{i}{\Gamma}{\emptyf[\Gamma]}{x}}{\jvtermeq{i}{\Gamma}{\emptyf[\Gamma]}{x}{\emptytm[\Gamma]}}
\end{align}

\begin{align}
& \inference{\jctx{\Gamma}}{\jvfam{\nameless}{\emptyc}{\Gamma}} 
& & \inference{\jvfam{\nameless}{\emptyc}{\Gamma}}{\jctx{\Gamma}}\\
& \inference{\jctxeq{\Gamma}{\Delta}}{\jvfameq{\nameless}{\emptyc}{\Gamma}{\Delta}}
& & \inference{\jvfameq{\nameless}{\emptyc}{\Gamma}{\Delta}}{\jctxeq{\Gamma}{\Delta}}
\end{align}

\subsubsection{The empty context is compatible with itslef}
The empty context $\emptyc$ may be considered as a family of contexts over the empty
context. When we do this, we get $\emptyf[\emptyc]$.
\begin{equation}
\inference{}{\jvfameq{\nameless}{\emptyc}{\emptyc}{\emptyf[\emptyc]}}
\end{equation}
In the future, we shall denote $\emptyf[\Gamma]$ by $\emptyf$. The above rule
guarantees that this will not cause confusion. Likewise, we shall denote
$\emptytm[\Gamma]$ by $\emptytm$.
